---
title: 'Assignment 3: Principal Components'
author: "Sergi Vilardell"
subtitle: Mathematics for Big Data
output:
  pdf_document: null
  html_document:
    highlights: pygments
    toc: yes
---

The goal of these exercises is to build a function to compute
principal components analysis, this will be done step by step. At the end, `PCA()` function in `FactoMineR` library is
explored, and the resutls compared.

```{r, message=FALSE, warning=FALSE}
library(FactoMineR)
library(fBasics)
```

# 1 Preliminaries

1. Use the data `decathlon` in package `FactoMineR`:

```{r}
data(decathlon, package="FactoMineR")
data<-decathlon
```

2. Transform the variables involving race times, so that: $t \rightarrow \max -t$:  

```{r}
time.cols<- data[c(1,5,6,10)]
max.cols<-apply(time.cols,2,max)
transf<- sweep(time.cols,2,max.cols,FUN="-")
transf<- -transf
data[c(1,5,6,10)]<-transf
```

3. Separate basic data from supplementary data: variables 11, 12 and 13 are considered supplementary, numerical
or qualitative, and cases 39 to 41 are considered as supplementary too:

```{r}
dat<-data[-(39:41),-(11:13)] # basic data subet
ind.supp<- data[39:41,-(11:13)] # supplementary individua
var.supp<- data[-(39:41),11:12] # supplementary numerical variables
qual.supp<- as.factor(data[-(39:41),13]) # supplementary categorical variable
```

We have our data set, now we can begin performing PCA.

# 2 Steps in the function

We want to create a new function `pc()` to get principal componenents and all the rellevant matrices involved
in the interpretation.

## 2.1 Normalizing

4. Normalize (scale) the data, i.e., substract the mean and divide by the standard deviation by columns.

```{r}
n<-nrow(dat)
p<-ncol(dat)

centr<-apply(dat,2,mean)   # global center: mean vector of the basic subset
s<-apply(dat,2,sd)         # global st.dev: st. dev of the basic subset

dat.centr<-sweep(dat,2,centr,FUN="-")
dat.std <-sweep(dat.centr,2,s,FUN="/")

# ind.supp (if not null) must be scaled with respect to global center and st.dev
if (!is.null(ind.supp)){
ind.supp.centr<-sweep(ind.supp,2,centr,FUN="-")
ind.supp.std <-sweep(ind.supp.centr,2,s,FUN="/")}

# var.supp subset (if not null) must be scaled with respect its own center
if (!is.null(var.supp)){
var.supp.centr<-sweep(var.supp,2,apply(var.supp,2,mean),FUN="-")
var.supp.std <-sweep(var.supp.centr,2,apply(var.supp,2,sd),FUN="/")}

# Warning in treating the quali.supp variable !!
# 1st: get the column means for the factor groups
if (!is.null(qual.supp)){
dat.qual.supp<-lapply(split(data.frame(dat),qual.supp),colMeans) # its a list

# 2nd: convert to a data frame
dat.qual.supp<-data.frame(dat.qual.supp)

# 3rd: transpose to trate groups as individual cases
dat.qual.supp<-t(dat.qual.supp)

# 4th: center and scale using descriptives of basic data
qual.supp.centr<-sweep(dat.qual.supp,2,centr,FUN="-")
qual.supp.std<-sweep(qual.supp.centr,2,s,FUN="/")}

```

5. Getting the data subsets to whom principal components apply (factor $1/\sqrt{n-1}$ applies to obtain the covariance).


```{r}
corr<-TRUE # standardized data  
if (corr==T){ X<-dat.std*(1/sqrt(n-1))
if (!is.null(ind.supp)){X.ind.supp<-ind.supp.std*(1/sqrt(n-1))}
if (!is.null(var.supp)){X.var.supp<-var.supp.std*(1/sqrt(n-1))}
if (!is.null(qual.supp)){X.qual.supp<-qual.supp.std*(1/sqrt(n-1))}
Sigma<-t(as.matrix(X))%*%as.matrix(X)
Sinv<-diag(rep(1,p))} # "Sinv"" is the identity if corr=TRUE

#corr<-FALSE
if (corr==F) {X<-dat.centr*(1/sqrt(n-1))
if (!is.null(ind.supp)){X.ind.supp<-ind.supp.centr*(1/sqrt(n-1))}
if (!is.null(var.supp)){X.var.supp<-var.supp.centr*(1/sqrt(n-1))}
if (!is.null(qual.supp)){X.qual.supp<-qual.supp.centr*(1/sqrt(n-1))}
Sigma<-t(as.matrix(X))%*%as.matrix(X) 
Sinv<-diag((diag(Sigma))^(-(1/2)))}
```

6. Get $U$, $V$ and $D$ components in **SVD** of $X$, you can use the function `svd`.

```{r}
sing <- svd(X)
U <- sing$u
V <- sing$v
D <- diag(sing$d)
Lambda <- D^(2)
```

7. Compute relevant matrices for the variables (column analysis).


```{r}
# by columns: contribution of the variables into the components
var.contr<- 100*(V^2)
colSums(var.contr)
rownames(var.contr)<-colnames(dat)
colnames(var.contr)<-paste("PC",1:ncol(var.contr),sep="")
var.contr


# correlations between variables and components
C <- Sinv%*%V%*%D
rownames(C)<-colnames(dat)
colnames(C)<-paste("PC",1:ncol(var.contr),sep="")

# coordinates of the initial variables
var.coord <- V%*%D
rownames(var.coord)<-colnames(dat)
colnames(var.coord)<-paste("PC",1:ncol(var.contr),sep="")

# squared cosines
var.cos2 <- C^2
rownames(var.cos2)<-colnames(dat)
colnames(var.cos2)<-paste("PC",1:ncol(var.contr),sep="")
```

8. Compute the relevant matrices for the individuals, as in 7.

```{r}
#Scores
Y <- as.matrix(X)%*%V*sqrt(n)
colnames(Y)<-paste("PC",1:ncol(var.contr),sep="")

#Contribution of the individuals into the components
sqrt.scores <- Y^2
sum <- colSums(sqrt.scores)
ind.contr <- sweep(sqrt.scores, 2, sum,FUN="/")
ind.contr <- 100*ind.contr
colSums(ind.contr)

#Individuals inertias
ind.iner <- rowSums(sqrt.scores)
ind.dist <- sqrt(ind.iner)

#Individuals squared cosines
ind.cos2 <- sweep(sqrt.scores, 2, ind.iner,FUN="/")
```

9. Make a `list` with all the results above
computed, using a reduced number of digits $d$, for instance $d = 3$. Add to the list result the coordinates for the supplementary objects:

```{r}
# assume k=3
var.supp.coord <- as.matrix(t(X.var.supp))%*%U
qual.supp.coord <- as.matrix(X.qual.supp)%*%V*sqrt(n)
ind.supp.coord <- as.matrix(X.ind.supp)%*%V*sqrt(n)

result<-list(left.sing.vec =  round(U, digits = 3), 
             right.sing.vec = round(V, digits = 3), 
             diag.sing.vec = round(D, digits = 3), 
             lambda. = round(Lambda, digits = 3), 
             diag.sing = round(D, digits = 3), 
             var.contr = round(var.contr, digits = 3), 
             corr = round(C, digits = 3), 
             var.coord = round(var.coord, digits = 3), 
             var.cos2 = round(var.cos2, digits = 3), 
             ind.coord = round(Y, digits = 3), 
             ind.contr = round(ind.contr, digits = 3), 
             ind.iner= round(ind.iner, digits = 3), 
             ind.dist = round(ind.dist, digits = 3), 
             ind.cos2 = round(ind.cos2, digits = 3),
             qual.supp.coord = round(qual.supp.coord, digits = 3), 
             ind.supp.coord = round(ind.supp.coord, digits = 3))
```

# The function `pc()`

11. Write the complete code for the function.

```{r}
pc<-function(dat,ind.supp=NULL,var.supp=NULL,qual.supp=NULL,corr=TRUE,k=3) #default
{
data(decathlon, package="FactoMineR")
data<-decathlon

time.cols<- data[c(1,5,6,10)]
max.cols<-apply(time.cols,2,max)
transf<- sweep(time.cols,2,max.cols,FUN="-")
transf<- -transf
data[c(1,5,6,10)]<-transf

n<-nrow(dat)
p<-ncol(dat)

centr<-apply(dat,2,mean)   # global center: mean vector of the basic subset
s<-apply(dat,2,sd)         # global st.dev: st. dev of the basic subset

dat.centr<-sweep(dat,2,centr,FUN="-")
dat.std <-sweep(dat.centr,2,s,FUN="/")

# ind.supp (if not null) must be scaled with respect to global center and st.dev
if (!is.null(ind.supp)){
ind.supp.centr<-sweep(ind.supp,2,centr,FUN="-")
ind.supp.std <-sweep(ind.supp.centr,2,s,FUN="/")}

# var.supp subset (if not null) must be scaled with respect its own center
if (!is.null(var.supp)){
var.supp.centr<-sweep(var.supp,2,apply(var.supp,2,mean),FUN="-")
var.supp.std <-sweep(var.supp.centr,2,apply(var.supp,2,sd),FUN="/")}

# Warning in treating the quali.supp variable !!
# 1st: get the column means for the factor groups
if (!is.null(qual.supp)){
dat.qual.supp<-lapply(split(data.frame(dat),qual.supp),colMeans) # its a list

# 2nd: convert to a data frame
dat.qual.supp<-data.frame(dat.qual.supp)

# 3rd: transpose to trate groups as individual cases
dat.qual.supp<-t(dat.qual.supp)

# 4th: center and scale using descriptives of basic data
qual.supp.centr<-sweep(dat.qual.supp,2,centr,FUN="-")
qual.supp.std<-sweep(qual.supp.centr,2,s,FUN="/")}

#corr<-TRUE 
if (corr==T){ X<-dat.std*(1/sqrt(n-1))
if (!is.null(ind.supp)){X.ind.supp<-ind.supp.std*(1/sqrt(n-1))}
if (!is.null(var.supp)){X.var.supp<-var.supp.std*(1/sqrt(n-1))}
if (!is.null(qual.supp)){X.qual.supp<-qual.supp.std*(1/sqrt(n-1))}
Sigma<-t(as.matrix(X))%*%as.matrix(X)
Sinv<-diag(rep(1,p))} # "Sinv"" is the identity if corr=TRUE

#corr<-FALSE
if (corr==F) {X<-dat.centr*(1/sqrt(n-1))
if (!is.null(ind.supp)){X.ind.supp<-ind.supp.centr*(1/sqrt(n-1))}
if (!is.null(var.supp)){X.var.supp<-var.supp.centr*(1/sqrt(n-1))}
if (!is.null(qual.supp)){X.qual.supp<-qual.supp.centr*(1/sqrt(n-1))}
Sigma<-t(as.matrix(X))%*%as.matrix(X)
Sinv<-diag((diag(Sigma))^(-(1/2)))}

#SVD
sing <- svd(X)
U <- sing$u
V <- sing$v
D <- diag(sing$d)
Lambda <- D^(2)

# by columns: contribution of the variables into the components
var.contr<- 100*(V^2)
colSums(var.contr)
rownames(var.contr)<-colnames(dat)
colnames(var.contr)<-paste("PC",1:ncol(var.contr),sep="")


# correlations between variables and components
C <- Sinv%*%V%*%D
rownames(C)<-colnames(dat)
colnames(C)<-paste("PC",1:ncol(var.contr),sep="")

# coordinates of the initial variables
var.coord <- V%*%D
rownames(var.coord)<-colnames(dat)
colnames(var.coord)<-paste("PC",1:ncol(var.contr),sep="")

# squared cosines
var.cos2 <- C^2
rownames(var.cos2)<-colnames(dat)
colnames(var.cos2)<-paste("PC",1:ncol(var.contr),sep="")

var.supp.coord <- as.matrix(t(X.var.supp))%*%U
qual.supp.coord <- as.matrix(X.qual.supp)%*%V*sqrt(n)
ind.supp.coord <- as.matrix(X.ind.supp)%*%V*sqrt(n)

result<-list(left.sing.vec =  round(U, digits = k), 
             right.sing.vec = round(V, digits = k), 
             diag.sing.vec = round(D, digits = k), 
             lambda.sing.vec = round(Lambda, digits = k), 
             var.contr = round(var.contr, digits = k), 
             corr = round(C, digits = k), 
             var.coord = round(var.coord, digits = k), 
             var.cos2 = round(var.cos2, digits = k), 
             ind.coord  = round(Y, digits = k), 
             ind.contr = round(ind.contr, digits = k), 
             ind.iner = round(ind.iner, digits = k), 
             ind.dist = round(ind.dist, digits = k), 
             ind.cos2 = round(ind.cos2, digits = k),
             qual.supp.coord = round(qual.supp.coord, digits = k), 
             ind.supp.coord = round(ind.supp.coord, digits = k),
             var.supp.coord = round(var.supp.coord, digits = k))
}
```


12. Apply function `pc()` in the following two cases and explore within objects `out` and `out2`.

```{r}
out<-pc(dat,ind.supp=NULL,var.supp=NULL,qual.supp=NULL,corr=TRUE)
out2<-pc(dat, ind.supp=ind.supp, var.supp=var.supp, qual.supp=qual.supp, corr=TRUE,k=8)
```

# 4 `PCA()` function in package `FactoMineR`

13. Compare the result in `out2` with the ones provided by the function `PCA()`.

```{r}
res<-PCA(data, ncp = 10, ind.sup = 39:41, quanti.sup = c(11: 12), quali.sup = 13,  graph=T)
```

We can compare some matrices of `pca()` and `PCA()` to see if they match. Some eigenvectors are multiplied by a factor $(-1)$. In order to compare them properly we take the absolute value of the matrices:

```{r}
#SVD
all.equal(abs(out2$right.sing.vec),abs(res$svd$V))

#Var. Coord. (Change the name of the columns, if not it gives an error...)
colnames(out2$var.coord)<-paste("Dim.",1:ncol(var.contr),sep="")
all.equal(abs(out2$var.coord),abs(res$var$coord))

#Ind. Coord.
colnames(out2$ind.coord)<-paste("Dim.",1:ncol(var.contr),sep="")
all.equal(abs(out2$ind.coord),abs(res$ind$coord))
```


14. Function `PCA()` has the default option graph=TRUE that displays 2 graphs. In particular, explore the arguments options: `axes`, `choix`, `ellipse`, `xlim`, `ylim`, `habillage`, `invisible`, `lim.cos2.var`,
`select`, etc. Use these options to explore other components 3 and 4, to represent the athletes in a diferent color depending on
`Decastar` or `OlymplicG`, etc.

```{r}
#Axes selects the components to plot
plot.PCA(res,axes = c(2,3))

#plot the variables
plot.PCA(res, choix = "var")

#plot ind. by group
plot.PCA(res, choix="ind", habillage=13)

#Make ind invisible
plot.PCA(res, choix="ind", habillage=13, invisible = "ind")

#Comp 3 and 4
plot.PCA(res,axes = c(3,4), habillage = 13)

```





