---
title: 'Assignment 4: Factor Analysis'
author: "Sergi Vilardell"
subtitle: Mathematics for Big Data
output:
  pdf_document: null
  html_document:
    highlights: pygments
    toc: yes
---

```{r}
library(psych)
library(datasets)
library(tidyverse)
```


# Maximum Likelihood method (MLE) in FA

MLE method to estimate the parameters $Q$ and $\Psi$ can be performed using function `factanal()`.

1.-Get the two-factor solution with the following options: `varimax rotation` and scores obtained
by the `regression` method. Save it in an object `ml.vx.reg2`.

(a) Obtain the loadings matrix (the complete one) and save it in object `Q`.

```{r}
#FA
data(LifeCycleSavings, package = "datasets")
data <- LifeCycleSavings
dat <- data.frame(data[, 1:4], logddpi = log(data$ddpi)) ## we use log(ddpi)
names(dat)<-c("sr","p15","p75","dpi","ldd") ## shorter names
head(dat,3)
fa.nr<-factanal(dat,factors=2,rotation="none",scores="regression")
fa.vx<-factanal(dat,factors=2,rotation="varimax",scores="regression")

```

```{r}
#Loadings matrix
Q <- fa.vx$loadings[,1:2]
Q.2 <- Q^2
hi2 <- data.frame(rowSums(Q.2))
colnames(hi2) <- "comm"
psi <- data.frame(apply(hi2,1, FUN = function(x) 1-x))
colnames(psi) <- "spec"
```


```{r}
#Scores matrix
F <- fa.vx$scores[,1:2]

#Biplot

biplot(Q,F)

```

```{r}
#New 2-factor promax

ml.px.reg2 <- factanal(dat,factors=2,rotation="promax",scores="regression")

Q.p <- ml.px.reg2$loadings[,1:2]
F.p <- ml.px.reg2$scores[,1:2]
biplot(Q,F)
```


#Principal Axes Factorial method (PAF) in FA

Apply PAF and MLE estimation methods, both with 4 factors and varimax rotation, to the $drug usage data$ using the function `fa()`. Compare the results provided by MLE and PAF.

```{r}
PAF <-  fa(druguse.cor, nfactors = 4, rotate = "varimax",fm = "pa")
MLE <-  fa(druguse.cor, nfactors = 4, rotate = "varimax",fm = "mle")
```

